---
title: "COVID-19 Analysis, track your county"
author: "LDL Anderegg"
date: "4/7/2020"
output: html_document
---

Figures made from NYT county-level and state-level aggregated data
Most recent data: 04/08/2020


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(ggplot2)
require(dplyr)
require(stringr)
require(RColorBrewer)
require(zoo)
mypal <- c(brewer.pal(n=9, "Set1"), brewer.pal(n=8, "Dark2"))
```

```{r import data}
  # Note: you need to have the NYT 'covid-19-data' directory in the same parent directory that you're storing the 'covid-analysis' directory that this file is in. otherwise, comment this bit out and input the desired path in the version right below it
county <- read.csv(paste0(gsub(pattern = "/covid-analysis",replacement="", getwd()),"/covid-19-data/us-counties.csv"))
# if specifying path
#county <- read.csv("~/Dropbox/covid-19-data/us-counties.csv")

county$date <- base::as.Date(county$date)
county$state.county <- paste(county$state, county$county, sep=".")

```

## Raw Cases

NYC dwarfs pretty much everywhere in raw number of cases. It goes waaaaay of the chart, but you can't see much even after I cut off the y axis


*Also pulling out all counties where a death was one of the first 20 cases reported, to see if they seem to be worse off then counties that either haven't reported a death or had >20 cases before a reported death. These counties will be in red from here on out*
``` {r find similar counties, include=F}
bad.counties <- unique(county$state.county[which(county$cases<= 10 & county$deaths>1)])

my.counties <- c("Colorado.Mesa", "California.Santa Clara", "Colorado.La Plata", "Oregon.Lane", "California.San Mateo", "Utah.San Juan", "New Mexico.San Juan", "New Mexico.Mckinley", "Arizona.Navajo","Arizona.Apache","Arizona.Coconino")
# plot(cases~date, county, pch=16, cex=.5, log="y")
# points(cases~date, county[grep("Montezuma", county$county), ], pch=3, col="red")
# legend("topleft", legend=c("all counties", "Montezuma County"), pch=c(16, 3), pt.cex=c(.5,1), col=c("black","red"))
# points(cases~date, county[which(county$county %in% bad.counties),], pch=16, cex=.6, col=county)
# 
# ggplot(county, aes(x=date, y=log(cases, base=10), col=state.county)) + geom_line() + theme(legend.position="none") + geom_point(data=county[grep("Montezuma", county$county),], col="black")
# 
# ggplot(county[which(county$cases>3 & !county$state.county %in% bad.counties),],  aes(x=date, y=log(cases, base=10), col=state.county)) + geom_line() + theme(legend.position="none")
# 
# ggplot(county[which(county$cases>3 & county$state.county %in% bad.counties),],  aes(x=date, y=log(cases, base=10), col=state.county)) + geom_line() + theme(legend.position="none")



```

```{r counties to track, echo=F}
print("We will track these counties in particular:")
print(paste(gsub(".*\\.","",my.counties), collapse=", "), font=2)
```


``` {r transform to date of first 3 cases, warning=FALSE, echo=F}
tmp <- county %>% group_by(state.county) %>% mutate(
  new.cases = c(NA,diff(cases, differences=1))
  , new.deaths = c(NA,diff(deaths, differences=1))
  , min.cases = min(cases[which(new.cases>0)], na.rm=T)
  , min.deaths=min(deaths[which(new.deaths>0)],na.rm=T)
  , first.case = min(date[which(new.cases>0)], na.rm=T)
  , first.death = min(date[which(new.deaths>0)], na.rm=T)
  , tenth.case = min(date[which(cases>=10)], na.rm=T)
  , tenth.death = min(date[which(deaths>=10)], na.rm=T)
  , max.cases=max(cases)
  , max.deaths=max(deaths)
  , new.cases.4avg =rollmean(new.cases, 4, fill=NA, align="right")
  , new.deaths.4avg = rollmean(new.deaths,4, fill=NA, align="right")) %>% arrange(state.county)
tmp$min.cases[which(tmp$min.cases==Inf)] <- NA
tmp$min.deaths[which(tmp$min.deaths==Inf)] <- NA
tmp$first.case[which(tmp$first.case==Inf)] <- NA
tmp$first.death[which(tmp$first.death==Inf)] <- NA

growing <- tmp[which(tmp$date>=tmp$first.case),]
growing$time.since.first <- growing$date - growing$first.case
growing$cases.since.min <- growing$cases - growing$min.cases
growing$time.since.tenth <- growing$date - growing$tenth.case
growing$time.since.tenth[which(growing$time.since.tenth<0)] <- NA

countysums <- growing[which(growing$max.cases>=10),] %>% group_by(state,state.county, county, max.cases, max.deaths) %>% summarise( time.to.50 = min(time.since.first[which(cases>= 50)]), time.to.10 = min(time.since.first[which(cases>= 10)]), time.to.10.deaths = min(time.since.first[which(deaths>= 10)]) )

countysums$early.deaths <- "no"
countysums$early.deaths[which(countysums$state.county %in% bad.counties)] <- "yes"
countysums$early.deaths <- factor(countysums$early.deaths)
countysums$time.to.10[which(countysums$time.to.10==Inf)] <- NA
countysums$time.to.50[which(countysums$time.to.50==Inf)] <- NA
countysums$time.to.10.deaths[which(countysums$time.to.10.deaths==Inf)] <- NA
# 
countysums %>% group_by(early.deaths) %>% summarise(mean.time.to.10 = mean(time.to.10, na.rm = T), median.time.to.10 = median(time.to.10, na.rm = T),mean.time.to.50 = mean(time.to.50, na.rm = T))
```

``` {r summary tables}


```


``` {r raw cases, basic time, warning=F, echo=F}

plot(cases~date, county, type="n", ylim=c(0,20000))
for(i in unique(county$state.county[which(!county$state.county %in% bad.counties)])){
  lines(cases~date, county[which(county$state.county==i),], col="#AAAAAA66")
}
for(i in unique(county$state.county[which(county$state.county %in% bad.counties)])){
  lines(cases~date, county[which(county$state.county==i),], col=paste0(mypal[1],"66"))
}

for(j in 1:length(my.counties)){
  lines(cases~date, growing[which(growing$state.county==my.counties[j]),], col=mypal[1+j], lwd=3)
}
points(cases~date, growing[which(growing$county=="Montezuma"),], pch=3, lwd=2, type="b")
legend('topleft', legend=c("all counties","early deaths","Montezuma",gsub(".*\\.","",my.counties)), lwd=c(1,1,2,rep(2, length(my.counties))), pch=c(NA,NA,3,NA,NA), col=c("#AAAAAA66", mypal[1],"black", mypal[2:(length(my.counties)+1)]))


```

## log-transformed cases
This squeezes the y-axis (to keep NYC from dominating) and illustrates the exponential growth of the virus (a fixed % growth-rate will be linear in this plot versus exponential in the previous plot)


``` {r log cases, basic time, warning=F, echo=F}

plot(cases~date, county, type="n", log="y")
for(i in unique(county$state.county[which(!county$state.county %in% bad.counties)])){
  lines(cases~date, county[which(county$state.county==i),], col="#AAAAAA66")
}
for(i in unique(county$state.county[which(county$state.county %in% bad.counties)])){
  lines(cases~date, county[which(county$state.county==i),], col=paste0(mypal[1],"66"))
}

for(j in 1:length(my.counties)){
  lines(cases~date, growing[which(growing$state.county==my.counties[j]),], col=mypal[1+j], lwd=3)
}
points(cases~date, growing[which(growing$county=="Montezuma"),], pch=3, lwd=2, type="b")
legend('topleft', legend=c("all counties","early deaths","Montezuma",gsub(".*\\.","",my.counties)), lwd=c(1,1,2,rep(2, length(my.counties))), pch=c(NA,NA,3,NA,NA), col=c("#AAAAAA66", mypal[1],"black", mypal[2:(length(my.counties)+1)]))




```


It looks like Montezuma County is growing rapidly in a relative sense (growing at ~100% pace every couple days), but hopefully that's just because numbers are small. 
-Santa Clara county (Stanford + San Jose) got out of the gates super early, but has been growing at a pretty compared to most other highly infected counties.
- La plata county jumped up to almost 30 cases super quickly but is pretty level thereafter. Maybe this is just a testing artifact (backlog cleared in late March, but now building another backlog?)
- Mesa County is growing fairly consistently, but still has >40 cases.

``` {r standardized by time.since.first, echo=F, warning=F}
plot(cases~time.since.first, growing, log="y", type="n", xlab="Time since first case")
# plot(log(cases~time.since.first, growing, type="n", xlab="Time since first case")
for(i in unique(growing$state.county[which(!growing$state.county %in% bad.counties)])){
  lines(cases~time.since.first, growing[which(growing$state.county==i),], col="#AAAAAA66")
}
for(i in unique(growing$state.county[which(growing$state.county %in% bad.counties)])){
  lines(cases~time.since.first, growing[which(growing$state.county==i),], col=paste0(mypal[1],"66"))
}
for(j in 1:length(my.counties)){
  lines(cases~time.since.first, growing[which(growing$state.county==my.counties[j]),], col=mypal[1+j], lwd=3)
}
points(cases~time.since.first, growing[which(growing$county=="Montezuma"),], pch=3, lwd=2, type="b")
legend('bottomright', legend=c("all counties","early deaths","Montezuma",gsub(".*\\.","",my.counties)), lwd=c(1,1,2,rep(2, length(my.counties))), pch=c(NA,NA,3,NA,NA), col=c("#AAAAAA66", mypal[1],"black", mypal[2:(length(my.counties)+1)]))
x <- 1:80
lines(cumsum(exp(.2333*x))~x, lty=2)
lines(cumsum(exp(.10*x))~x, lty=3)
legend("topright", legend=c("3 days", "7 days"), lty=c(2,3),title = "Doubling Time", horiz=T, bty="n")
# points(cases~time.since.first, growing[which(growing$state.county %in% bad.counties),], pch=16, col="darkred", cex=.5)
# points(cases~time.since.first, growing[grep("Montezuma", growing$county),], pch=15, col="red", cex=1)

```



If we standardize things by the time when the first case was reported, we can compare the slopes a bit better (except for Sant Clara, which sat at a few cases for almost a month). That other county out by Santa Clara must be King County around Seattle (also doing a pretty good job with slow growth)
-Thank god none of us seem to be growing like the real high infection hotspots. Doubling every week is still real scary, but better than doubling every 1-3 days, which is what NYC, New Orleans, and a few other places were doing for a while.

We can also see the 'exponential growth' trajectories by looking just at the number of new cases rather than total cases:


```{r the CURVE standardized by first case, echo=F, warning=F}
par(mar=c(4,4,1,1))
plot(new.cases.4avg~time.since.first, growing, type="n",log="y", ylab="4 day avg new cases", xlab="days since first case")
for(i in unique(growing$state.county[which(!growing$state.county %in% bad.counties)])){
  lines(new.cases.4avg~time.since.first, growing[which(growing$state.county==i),], col="#AAAAAA66")
}
for(i in unique(growing$state.county[which(growing$state.county %in% bad.counties)])){
  lines(new.cases.4avg~time.since.first, growing[which(growing$state.county==i),], col=paste0(mypal[1],"66"))
}
# plot counties of interest
for(j in 1:length(my.counties)){
  lines(new.cases.4avg~time.since.first, growing[which(growing$state.county==my.counties[j]),], col=mypal[1+j], lwd=3)
}
# Montezuma County
points(new.cases.4avg~time.since.first, growing[which(growing$county=="Montezuma"),], pch=3, lwd=2, type="b")

legend('bottomright', legend=c("all counties","early deaths","Montezuma",gsub(".*\\.","",my.counties)), lwd=c(1,1,2,rep(2, length(my.counties))), pch=c(NA,NA,3,NA,NA), col=c("#AAAAAA66", mypal[1],"black", mypal[2:(length(my.counties)+1)]), ncol=2, cex=.95)
x <- 1:80
lines(exp(.2333*x)~x, lty=2)
lines(exp(.10*x)~x, lty=3)
legend("topright", legend=c("3 days", "7 days"), lty=c(2,3),title = "Doubling Time", horiz=T, bty="n")




plot(new.cases.4avg~time.since.tenth, growing, type="n",log="y", ylab="4 day avg new cases", xlab="days since 10th case")
for(i in unique(growing$state.county[which(!growing$state.county %in% bad.counties)])){
  lines(new.cases.4avg~time.since.tenth, growing[which(growing$state.county==i),], col="#AAAAAA66")
}
for(i in unique(growing$state.county[which(growing$state.county %in% bad.counties)])){
  lines(new.cases.4avg~time.since.tenth, growing[which(growing$state.county==i),], col=paste0(mypal[1],"66"))
}
# plot counties of interest
for(j in 1:length(my.counties)){
  lines(new.cases.4avg~time.since.tenth, growing[which(growing$state.county==my.counties[j]),], col=mypal[1+j], lwd=3)
}
# Montezuma County
points(new.cases.4avg~time.since.tenth, growing[which(growing$county=="Montezuma"),], pch=3, lwd=2, type="b")

legend('bottomright', legend=c("all counties","early deaths","Montezuma",gsub(".*\\.","",my.counties)), lwd=c(1,1,2,rep(2, length(my.counties))), pch=c(NA,NA,3,NA,NA), col=c("#AAAAAA66", mypal[1],"black", mypal[2:(length(my.counties)+1)]), ncol=2)
x <- 1:80
lines(exp(.2333*x)~x, lty=2)
lines(exp(.10*x)~x, lty=3)
legend("topright", legend=c("3 days", "7 days"), lty=c(2,3),title = "Doubling Time", horiz=T, bty="n")

```



Same curves but standardized to plot versus days since first case or days since 10th case. a few counties look like they're peaking? But really hard to say. Let's look at a different way of assessing this:




## 'THE CURVE' 
The epidemiological 'curve' of new cases that you've no doubt seen everywhere, the thing we're trying to flatten

``` {r curves, warning=FALSE, echo=FALSE}

plot(new.cases.4avg~date, growing, type="n",log="y", ylab="4 day avg new cases")
for(i in unique(growing$state.county[which(!growing$state.county %in% bad.counties)])){
  lines(new.cases.4avg~date, growing[which(growing$state.county==i),], col="#AAAAAA66")
}
for(i in unique(growing$state.county[which(growing$state.county %in% bad.counties)])){
  lines(new.cases.4avg~date, growing[which(growing$state.county==i),], col=paste0(mypal[1],"66"))
}
# plot counties of interest
for(j in 1:length(my.counties)){
  lines(new.cases.4avg~date, growing[which(growing$state.county==my.counties[j]),], col=mypal[1+j], lwd=3)
}
# Montezuma County
points(new.cases.4avg~date, growing[which(growing$county=="Montezuma"),], pch=3, lwd=2, type="b")

legend('topleft', legend=c("all counties","early deaths","Montezuma",gsub(".*\\.","",my.counties)), lwd=c(1,1,2,rep(2, length(my.counties))), pch=c(NA,NA,3,NA,NA), col=c("#AAAAAA66", mypal[1],"black", mypal[2:(length(my.counties)+1)]))
```


This is real noisy, because in most counties we don't have a ton of new cases per day yet. But one could potentially take heart on the shape of some of these curves. Things *might* be going down (though the accuracy of recent testing data is highly dubious until we're a few days out due to backlogs and such).



**Better 'Curve' assessment?**

``` {r cumulative plot, warning=F, echo=F}


plot(new.cases.4avg~cases, growing, type="n", log="xy",xlab="total cases", ylab="new cases (4 day avg)", yaxt="n", xaxt="n")
axis(2, at = c(1,10,100,1000), labels = c(1,10,100,1000))
axis(1, at= c(1,2,10,20,100,200,1000,2000,10000), labels=c(1,2,10,20,100,200,1000,2000,10000))
for(i in unique(growing$state.county[which(!growing$state.county %in% bad.counties)])){
  lines(new.cases.4avg~cases, growing[which(growing$state.county==i),], col="#AAAAAA66")
}
for(i in unique(growing$state.county[which(growing$state.county %in% bad.counties)])){
  lines(new.cases.4avg~cases, growing[which(growing$state.county==i),], col=paste0(mypal[1],"66"))
}
# plot counties of interest
for(j in 1:length(my.counties)){
  lines(new.cases.4avg~cases, growing[which(growing$state.county==my.counties[j]),], col=mypal[1+j], lwd=3)
}
# Montezuma County
points(new.cases.4avg~cases, growing[which(growing$county=="Montezuma"),], pch=3, lwd=2, type="b")

legend('bottomright', legend=c("all counties","early deaths","Montezuma",gsub(".*\\.","",my.counties)), lwd=c(1,1,2,rep(2, length(my.counties))), pch=c(NA,NA,3,NA,NA), col=c("#AAAAAA66", mypal[1],"black", mypal[2:(length(my.counties)+1)]))
```

This is another way of looking to see when exponential growth gets shut down. X axis is total number of cases, y axis is new cases per day (averaged over past 4 days to get rid of some noise, you can still see weekend slumps in test returns though). When new cases stop being a funciton of total cases (i.e. exponential growth stops), the lines will fall off the diagonal. Unfortunately, no one is really doing that yet. 

```{r deaths vs cummulative deaths, echo=F, warning=F}
### Plotting deaths instead of cases
plot(new.deaths.4avg~deaths, growing, type="n", log="xy",xlab="total deaths", ylab="new deaths (4 day avg)")
#axis(2, at = c(1,10,100,1000), labels = c(1,10,100,1000))
#axis(1, at= c(1,2,10,20,100,200,1000,2000,10000), labels=c(1,2,10,20,100,200,1000,2000,10000))
for(i in unique(growing$state.county[which(!growing$state.county %in% bad.counties)])){
  lines(new.deaths.4avg~deaths, growing[which(growing$state.county==i),], col="#AAAAAA66")
}
for(i in unique(growing$state.county[which(growing$state.county %in% bad.counties)])){
  lines(new.deaths.4avg~deaths, growing[which(growing$state.county==i),], col=paste0(mypal[1],"66"))
}
# plot counties of interest
for(j in 1:length(my.counties)){
  lines(new.deaths.4avg~deaths, growing[which(growing$state.county==my.counties[j]),], col=mypal[1+j], lwd=3)
}
# Montezuma County
points(new.deaths.4avg~deaths, growing[which(growing$county=="Montezuma"),], pch=3, lwd=2, type="b")

legend('bottomright', legend=c("all counties","early deaths","Montezuma",gsub(".*\\.","",my.counties)), lwd=c(1,1,2,rep(2, length(my.counties))), pch=c(NA,NA,3,NA,NA), col=c("#AAAAAA66", mypal[1],"black", mypal[2:(length(my.counties)+1)]))
```

Doing the same thing for deaths rather than cases, we see that, thankfully, we don't have enough county-level deaths for this to work in most cases.

*****

------
### TRACK YOUR STATE
And last but not least, we can look at state totals just for fun (though the NYT does a better job visualizing this)
```{r states, include=F}
my.states <- c("California","Utah","Colorado", "Oregon","Florida")
state <- read.csv(paste0(gsub(pattern = "/covid-analysis",replacement="", getwd()),"/covid-19-data/us-states.csv"))
state$date <- base::as.Date(state$date)
tmp <- state %>% group_by(state) %>% mutate(
  new.cases = c(NA,diff(cases, differences=1))
  , new.deaths = c(NA,diff(deaths, differences=1))
  , min.cases = min(cases[which(new.cases>0)], na.rm=T)
  , min.deaths=min(deaths[which(new.deaths>0)],na.rm=T)
  , first.case = min(date[which(new.cases>0)], na.rm=T)
  , first.death = min(date[which(new.deaths>0)], na.rm=T)
  , tenth.case = min(date[which(cases>=10)], na.rm=T)
  , tenth.death = min(date[which(deaths>=10)], na.rm=T)
  , max.cases=max(cases)
  , max.deaths=max(deaths)
  , new.cases.4avg =rollmean(new.cases, 4, fill=NA, align="right")
  , new.deaths.4avg = rollmean(new.deaths,4, fill=NA, align="right")) %>% arrange(state)
tmp$min.cases[which(tmp$min.cases==Inf)] <- NA
tmp$min.deaths[which(tmp$min.deaths==Inf)] <- NA
tmp$first.case[which(tmp$first.case==Inf)] <- NA
tmp$first.death[which(tmp$first.death==Inf)] <- NA

state.growing <- tmp[which(tmp$date>=tmp$first.case),]
state.growing$time.since.first <- state.growing$date - state.growing$first.case
state.growing$cases.since.min <- state.growing$cases - state.growing$min.cases
state.growing$time.since.tenth <- state.growing$date - state.growing$tenth.case
state.growing$time.since.tenth[which(state.growing$time.since.tenth<0)] <- NA

statesums <- state.growing[which(state.growing$max.cases>=10),] %>% group_by(state, max.cases, max.deaths) %>% summarise( time.to.50 = min(time.since.first[which(cases>= 50)]), time.to.10 = min(time.since.first[which(cases>= 10)]), time.to.10.deaths = min(time.since.first[which(deaths>= 10)]) )

```


``` {r state curves, warning=FALSE, echo=FALSE}

plot(new.cases.4avg~date, state.growing, type="n",log="y", ylab="4 day avg new cases", main="New Cases over time")
for(i in unique(state.growing$state)){
  lines(new.cases.4avg~date, state.growing[which(state.growing$state==i),], col="#AAAAAA66")
}

# plot counties of interest
for(j in 1:length(my.states)){
  lines(new.cases.4avg~date, state.growing[which(state.growing$state==my.states[j]),], col=mypal[1+j], lwd=3)
}

legend('topleft', legend=c("all states",gsub(".*\\.","",my.states)), lwd=c(1,rep(2, length(my.states))), col=c("#AAAAAA66", mypal[2:(length(my.states)+1)]))
```


**Same curves, but standardizing the time axis to put all counties on the same timeline:**

```{r state the CURVE standardized by first case, echo=F, warning=F}
par(mar=c(4,4,1,1))

plot(new.cases.4avg~time.since.first, state.growing, type="n",log="y", ylab="4 day avg new cases", main="New Cases since First Case")
for(i in unique(state.growing$state)){
  lines(new.cases.4avg~time.since.first, state.growing[which(state.growing$state==i),], col="#AAAAAA66")
}

# plot counties of interest
for(j in 1:length(my.states)){
  lines(new.cases.4avg~time.since.first, state.growing[which(state.growing$state==my.states[j]),], col=mypal[1+j], lwd=3)
}

legend('topleft', legend=c("all states",gsub(".*\\.","",my.states)), lwd=c(1,rep(2, length(my.states))), col=c("#AAAAAA66", mypal[2:(length(my.states)+1)]))




plot(new.cases.4avg~time.since.tenth, state.growing, type="n",log="y", ylab="4 day avg new cases", main="Total Cases since 10th Case")
for(i in unique(state.growing$state)){
  lines(new.cases.4avg~time.since.tenth, state.growing[which(state.growing$state==i),], col="#AAAAAA66")
}

# plot counties of interest
for(j in 1:length(my.states)){
  lines(new.cases.4avg~time.since.tenth, state.growing[which(state.growing$state==my.states[j]),], col=mypal[1+j], lwd=3)
}

legend('topleft', legend=c("all states",gsub(".*\\.","",my.states)), lwd=c(1,rep(2, length(my.states))), col=c("#AAAAAA66", mypal[2:(length(my.states)+1)]))

```

**New Cases vs Total Cases**

``` {r state cumulative plot, warning=F, echo=F}
plot(new.deaths.4avg~deaths, state.growing, type="n",log="xy", ylab="4 day avg new deaths")
for(i in unique(state.growing$state)){
  lines(new.deaths.4avg~deaths, state.growing[which(state.growing$state==i),], col="#AAAAAA66")
}

# plot counties of interest
for(j in 1:length(my.states)){
  lines(new.deaths.4avg~deaths, state.growing[which(state.growing$state==my.states[j]),], col=mypal[1+j], lwd=3)
}

legend('topleft', legend=c("all states",gsub(".*\\.","",my.states)), lwd=c(1,rep(2, length(my.states))), col=c("#AAAAAA66", mypal[2:(length(my.states)+1)]))

```


No states really seem to be peaking and falling off of the exponential growth phase yet... 







``` {r totals}
plot(max.cases~max.deaths, countysums[which(countysums$state %in% c("Colorado","Utah","New Mexico", "Arizona")),], pch=16, col=state)

tmp <- countysums[which(countysums$state %in% c("Colorado","Utah","New Mexico", "Arizona")),] %>% arrange(max.cases)
tmp[which(tmp$state.county %in% my.counties),]
```
