---
title: "COVID-19 Analysis, track your county"
author: "LDL Anderegg"
date: "4/7/2020"
output: html_document
---

Figures made from NYT county-level and state-level aggregated data
Data repository: https://github.com/nytimes/covid-19-data

to pull from master: git pull upstream master

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(ggplot2)
require(dplyr)
require(stringr)
require(RColorBrewer)
require(zoo)
library(knitr)
mypal <- c(brewer.pal(n=9, "Set1"), brewer.pal(n=8, "Dark2"))
```

```{r import data, echo=F}
  # Note: you need to have the NYT 'covid-19-data' directory in the same parent directory that you're storing the 'covid-analysis' directory that this file is in. otherwise, comment this bit out and input the desired path in the version right below it
county <- read.csv(paste0(gsub(pattern = "/covid-analysis",replacement="", getwd()),"/covid-19-data/us-counties.csv"))
# if specifying path
#county <- read.csv("~/Dropbox/covid-19-data/us-counties.csv")

county$date <- base::as.Date(county$date)
county$state.county <- paste(county$state, county$county, sep=".")

```
Most recent data: 
```{r date}
print(max(county$date))
```

## Raw Cases

NYC dwarfs pretty much everywhere in raw number of cases. It goes waaaaay of the chart, but you can't see much even after I cut off the y axis


*Also pulling out all counties where a death was one of the first 20 cases reported, to see if they seem to be worse off then counties that either haven't reported a death or had >20 cases before a reported death. These counties will be in red from here on out*
``` {r find similar counties, include=F}
bad.counties <- unique(county$state.county[which(county$cases<= 10 & county$deaths>1)])

my.counties <- c("Colorado.Mesa", "California.Santa Clara", "Colorado.La Plata", "Oregon.Lane","Utah.Salt Lake", "California.Alameda", "New Mexico.San Juan", "Arizona.Navajo", "Colorado.Dolores")
all.my.counties <- c("Colorado.Montezuma","Colorado.Mesa","Colorado.La Plata", "Oregon.Lane", "Arizona.Maricopa","California.Santa Clara", "California.San Mateo","California.Alemeda","California.Contra Costa", "Utah.Salt Lake", "Utah.San Juan", "New Mexico.San Juan", "New Mexico.Mckinley", "Arizona.Navajo","Arizona.Apache","Arizona.Coconino", "Colorado.Conejos")
ca_counties <- c("California.Santa Clara","California.Alameda","California.San Mateo", "California.Contra Costa")
or_counties <- c("Oregon.Lane", "Oregon.Linn", "Oregon.Multnomah")
co_counties <- c("Colorado.Montezuma","Colorado.Mesa","Colorado.La Plata","New Mexico.San Juan", "Arazona.Navajo")
ut_counties <- c("Utah.Salt Lake")
# plot(cases~date, county, pch=16, cex=.5, log="y")
# points(cases~date, county[grep("Montezuma", county$county), ], pch=3, col="red")
# legend("topleft", legend=c("all counties", "Montezuma County"), pch=c(16, 3), pt.cex=c(.5,1), col=c("black","red"))
# points(cases~date, county[which(county$county %in% bad.counties),], pch=16, cex=.6, col=county)
# 
# ggplot(county, aes(x=date, y=log(cases, base=10), col=state.county)) + geom_line() + theme(legend.position="none") + geom_point(data=county[grep("Montezuma", county$county),], col="black")
# 
# ggplot(county[which(county$cases>3 & !county$state.county %in% bad.counties),],  aes(x=date, y=log(cases, base=10), col=state.county)) + geom_line() + theme(legend.position="none")
# 
# ggplot(county[which(county$cases>3 & county$state.county %in% bad.counties),],  aes(x=date, y=log(cases, base=10), col=state.county)) + geom_line() + theme(legend.position="none")



```

```{r counties to track, echo=F}
print("We will track these counties in particular:")
print(paste(gsub(".*\\.","",my.counties), collapse=", "), font=2)
```


``` {r transform to date of first 3 cases, warning=FALSE, echo=F}
tmp <- county %>% group_by(state.county) %>% mutate(
  new.cases = c(NA,diff(cases, differences=1))
  , new.deaths = c(NA,diff(deaths, differences=1))
  , min.cases = min(cases[which(new.cases>0)], na.rm=T)
  , min.deaths=min(deaths[which(new.deaths>0)],na.rm=T)
  , first.case = min(date[which(new.cases>0)], na.rm=T)
  , first.death = min(date[which(new.deaths>0)], na.rm=T)
  , tenth.case = min(date[which(cases>=10)], na.rm=T)
  , tenth.death = min(date[which(deaths>=10)], na.rm=T)
  , max.cases=max(cases)
  , max.deaths=max(deaths)
  , new.cases.4avg =rollmean(new.cases, 4, fill=NA, align="right")
  , new.deaths.4avg = rollmean(new.deaths,4, fill=NA, align="right")) %>% arrange(state.county)
tmp$min.cases[which(tmp$min.cases==Inf)] <- NA
tmp$min.deaths[which(tmp$min.deaths==Inf)] <- NA
tmp$first.case[which(tmp$first.case==Inf)] <- NA
tmp$first.death[which(tmp$first.death==Inf)] <- NA

growing <- tmp[which(tmp$date>=tmp$first.case),]
growing$time.since.first <- growing$date - growing$first.case
growing$cases.since.min <- growing$cases - growing$min.cases
growing$time.since.tenth <- growing$date - growing$tenth.case
growing$time.since.tenth[which(growing$time.since.tenth<0)] <- NA

countysums <- growing[which(growing$max.cases>=10),] %>% group_by(state,state.county, county, max.cases, max.deaths) %>% summarise( time.to.50 = min(time.since.first[which(cases>= 50)]), time.to.10 = min(time.since.first[which(cases>= 10)]), time.to.10.deaths = min(time.since.first[which(deaths>= 10)]) )

countysums$early.deaths <- "no"
countysums$early.deaths[which(countysums$state.county %in% bad.counties)] <- "yes"
countysums$early.deaths <- factor(countysums$early.deaths)
countysums$time.to.10[which(countysums$time.to.10==Inf)] <- NA
countysums$time.to.50[which(countysums$time.to.50==Inf)] <- NA
countysums$time.to.10.deaths[which(countysums$time.to.10.deaths==Inf)] <- NA
# # 
# countysums %>% group_by(early.deaths) %>% summarise(mean.time.to.10 = mean(time.to.10, na.rm = T), median.time.to.10 = median(time.to.10, na.rm = T),mean.time.to.50 = mean(time.to.50, na.rm = T))
```

``` {r summary tables, echo=F, warning=F}
tmp.for.sums <- growing[which(growing$state.county %in% all.my.counties & growing$date > (max(growing$date)- 7)),]

my.county.sums <- tmp.for.sums %>% group_by(state, county) %>% summarise(cases = max(cases), deaths = max(deaths), avg.case.perday = round(mean(new.cases.4avg, na.rm=T),1), avg.death.perday=round(mean(new.deaths.4avg, na.rm=T),1), time.since.first.case = max(time.since.first, na.rm=T))
# kable(my.county.sums)
# growth.rate <- rep(NA, length(unique(tmp.for.sums$state.county)))
# for(i in 1:length(unique(tmp.for.sums$state.county))){
#   tmp.count <- tmp.for.sums[which(tmp.for.sums$state.county==unique(tmp.for.sums$state.county)[i]),]
#   growth.rate[i] <- round(lm(cases~date, tmp.count)$coefficients["date"],1)
# }



```


``` {r raw cases, basic time, warning=F, echo=F}

plot(cases~date, county, type="n", ylim=c(0,20000))
for(i in unique(county$state.county[which(!county$state.county %in% bad.counties)])){
  lines(cases~date, county[which(county$state.county==i),], col="#AAAAAA66")
}
for(i in unique(county$state.county[which(county$state.county %in% bad.counties)])){
  lines(cases~date, county[which(county$state.county==i),], col=paste0(mypal[1],"66"))
}

for(j in 1:length(my.counties)){
  lines(cases~date, growing[which(growing$state.county==my.counties[j]),], col=mypal[1+j], lwd=3)
}
points(cases~date, growing[which(growing$county=="Montezuma"),], pch=3, lwd=2, type="b")
legend('topleft', legend=c("all counties","early deaths","Montezuma",gsub(".*\\.","",my.counties)), lwd=c(1,1,2,rep(2, length(my.counties))), pch=c(NA,NA,3,NA,NA), col=c("#AAAAAA66", mypal[1],"black", mypal[2:(length(my.counties)+1)]))


```

## log-transformed cases
This squeezes the y-axis (to keep NYC from dominating) and illustrates the exponential growth of the virus (a fixed % growth-rate will be linear in this plot versus exponential in the previous plot)


``` {r log cases, basic time, warning=F, echo=F}

plot(cases~date, county, type="n", log="y")
for(i in unique(county$state.county[which(!county$state.county %in% bad.counties)])){
  lines(cases~date, county[which(county$state.county==i),], col="#AAAAAA66")
}
for(i in unique(county$state.county[which(county$state.county %in% bad.counties)])){
  lines(cases~date, county[which(county$state.county==i),], col=paste0(mypal[1],"66"))
}

for(j in 1:length(my.counties)){
  lines(cases~date, growing[which(growing$state.county==my.counties[j]),], col=mypal[1+j], lwd=3)
}
points(cases~date, growing[which(growing$county=="Montezuma"),], pch=3, lwd=2, type="b")
legend('topleft', legend=c("all counties","early deaths","Montezuma",gsub(".*\\.","",my.counties)), lwd=c(1,1,2,rep(2, length(my.counties))), pch=c(NA,NA,3,NA,NA), col=c("#AAAAAA66", mypal[1],"black", mypal[2:(length(my.counties)+1)]))




```


It looks like Montezuma County is growing rapidly in a relative sense (growing at ~100% pace every couple days), but hopefully that's just because numbers are small. 
-Santa Clara county (Stanford + San Jose) got out of the gates super early, but has been growing at a pretty compared to most other highly infected counties.
- La plata county jumped up to almost 30 cases super quickly but is pretty level thereafter. Maybe this is just a testing artifact (backlog cleared in late March, but now building another backlog?)
- Mesa County is growing fairly consistently, but still has >40 cases.

``` {r standardized by time.since.first, echo=F, warning=F}
plot(cases~time.since.first, growing, log="y", type="n", xlab="Time since first case")
# plot(log(cases~time.since.first, growing, type="n", xlab="Time since first case")
for(i in unique(growing$state.county[which(!growing$state.county %in% bad.counties)])){
  lines(cases~time.since.first, growing[which(growing$state.county==i),], col="#AAAAAA66")
}
for(i in unique(growing$state.county[which(growing$state.county %in% bad.counties)])){
  lines(cases~time.since.first, growing[which(growing$state.county==i),], col=paste0(mypal[1],"66"))
}
for(j in 1:length(my.counties)){
  lines(cases~time.since.first, growing[which(growing$state.county==my.counties[j]),], col=mypal[1+j], lwd=3)
}
points(cases~time.since.first, growing[which(growing$county=="Montezuma"),], pch=3, lwd=2, type="b")
legend('bottomright', legend=c("all counties","early deaths","Montezuma",gsub(".*\\.","",my.counties)), lwd=c(1,1,2,rep(2, length(my.counties))), pch=c(NA,NA,3,NA,NA), col=c("#AAAAAA66", mypal[1],"black", mypal[2:(length(my.counties)+1)]))
x <- 1:80
lines(cumsum(exp(.2333*x))~x, lty=2)
lines(cumsum(exp(.10*x))~x, lty=3)
legend("topright", legend=c("3 days", "7 days"), lty=c(2,3),title = "Doubling Time", horiz=T, bty="n")
# points(cases~time.since.first, growing[which(growing$state.county %in% bad.counties),], pch=16, col="darkred", cex=.5)
# points(cases~time.since.first, growing[grep("Montezuma", growing$county),], pch=15, col="red", cex=1)

```



If we standardize things by the time when the first case was reported, we can compare the slopes a bit better (except for Sant Clara, which sat at a few cases for almost a month). That other county out by Santa Clara must be King County around Seattle (also doing a pretty good job with slow growth)
-Thank god none of us seem to be growing like the real high infection hotspots. Doubling every week is still real scary, but better than doubling every 1-3 days, which is what NYC, New Orleans, and a few other places were doing for a while.

We can also see the 'exponential growth' trajectories by looking just at the number of new cases rather than total cases:


```{r the CURVE standardized by first case, echo=F, warning=FALSE}
par(mar=c(4,4,1,1))
plot(new.cases.4avg~time.since.first, growing, type="n",log="y", ylab="4 day avg new cases", xlab="days since first case")
for(i in unique(growing$state.county[which(!growing$state.county %in% bad.counties)])){
  lines(new.cases.4avg~time.since.first, growing[which(growing$state.county==i),], col="#AAAAAA66")
}
for(i in unique(growing$state.county[which(growing$state.county %in% bad.counties)])){
  lines(new.cases.4avg~time.since.first, growing[which(growing$state.county==i),], col=paste0(mypal[1],"66"))
}
# plot counties of interest
for(j in 1:length(my.counties)){
  lines(new.cases.4avg~time.since.first, growing[which(growing$state.county==my.counties[j]),], col=mypal[1+j], lwd=3)
}
# Montezuma County
points(new.cases.4avg~time.since.first, growing[which(growing$county=="Montezuma"),], pch=3, lwd=2, type="b")

legend('bottomright', legend=c("all counties","early deaths","Montezuma",gsub(".*\\.","",my.counties)), lwd=c(1,1,2,rep(2, length(my.counties))), pch=c(NA,NA,3,NA,NA), col=c("#AAAAAA66", mypal[1],"black", mypal[2:(length(my.counties)+1)]), ncol=2, cex=.95)
x <- 1:80
lines(exp(.2333*x)~x, lty=2)
lines(exp(.10*x)~x, lty=3)
legend("topright", legend=c("3 days", "7 days"), lty=c(2,3),title = "Doubling Time", horiz=T, bty="n")




plot(new.cases.4avg~time.since.tenth, growing, type="n",log="y", ylab="4 day avg new cases", xlab="days since 10th case")
for(i in unique(growing$state.county[which(!growing$state.county %in% bad.counties)])){
  lines(new.cases.4avg~time.since.tenth, growing[which(growing$state.county==i),], col="#AAAAAA66")
}
for(i in unique(growing$state.county[which(growing$state.county %in% bad.counties)])){
  lines(new.cases.4avg~time.since.tenth, growing[which(growing$state.county==i),], col=paste0(mypal[1],"66"))
}
# plot counties of interest
for(j in 1:length(my.counties)){
  lines(new.cases.4avg~time.since.tenth, growing[which(growing$state.county==my.counties[j]),], col=mypal[1+j], lwd=3)
}
# Montezuma County
points(new.cases.4avg~time.since.tenth, growing[which(growing$county=="Montezuma"),], pch=3, lwd=2, type="b")

legend('bottomright', legend=c("all counties","early deaths","Montezuma",gsub(".*\\.","",my.counties)), lwd=c(1,1,2,rep(2, length(my.counties))), pch=c(NA,NA,3,NA,NA), col=c("#AAAAAA66", mypal[1],"black", mypal[2:(length(my.counties)+1)]), ncol=2)
x <- 1:80
lines(exp(.2333*x)~x, lty=2)
lines(exp(.10*x)~x, lty=3)
legend("topright", legend=c("3 days", "7 days"), lty=c(2,3),title = "Doubling Time", horiz=T, bty="n")




par(mfrow=c(1,2))
p <- plot(new.cases.4avg~time.since.tenth, growing[grep("Colorado",growing$state),], type="n",log="y", ylab="4 day avg new cases", xlab="days since 10th case", main="Colorado")
for(i in unique(growing$state.county[grep("Colorado",growing$state)])){
  lines(new.cases.4avg~time.since.tenth, growing[which(growing$state.county==i),], col="#AAAAAA66")
}

# plot counties of interest
for(j in 1:length(co_counties)){
  lines(new.cases.4avg~time.since.tenth, growing[which(growing$state.county==co_counties[j]),], col=mypal[j], lwd=3)
}
# Montezuma County
legend(x=par("usr")[3], y= par("usr")[2] + par("usr")[2], legend=c(gsub(".*\\.","",co_counties)), lwd=2, col=c(mypal[1:(length(my.counties)+1)]), ncol=length(co_counties)/2, xpd=NA)
x <- 1:80
lines(exp(.2333*x)~x, lty=2)
lines(exp(.10*x)~x, lty=3)
legend("topright", legend=c("3 days", "7 days"), lty=c(2,3),title = "Doubling Time", horiz=T, bty="n")



```



Same curves but standardized to plot versus days since first case or days since 10th case. a few counties look like they're peaking? But really hard to say. Let's look at a different way of assessing this:

***Looking at raw (unlogged) growth rates*** 
All of our coutnies (except perhaps Maricopa) seem to be doing better than many. Growth rates still need to go DOWN before things can get better, but we've evidently escaped the most viscious exponential growth (e.g. following the 'doubling ever 3 days' trajectory)

```{r the CURVE standardized by first case unlogged, echo=F, warning=FALSE}
# par(mar=c(4,4,1,1))
# plot(new.cases.4avg~time.since.first, growing, type="n",log="y", ylab="4 day avg new cases", xlab="days since first case")
# for(i in unique(growing$state.county[which(!growing$state.county %in% bad.counties)])){
#   lines(new.cases.4avg~time.since.first, growing[which(growing$state.county==i),], col="#AAAAAA66")
# }
# for(i in unique(growing$state.county[which(growing$state.county %in% bad.counties)])){
#   lines(new.cases.4avg~time.since.first, growing[which(growing$state.county==i),], col=paste0(mypal[1],"66"))
# }
# # plot counties of interest
# for(j in 1:length(my.counties)){
#   lines(new.cases.4avg~time.since.first, growing[which(growing$state.county==my.counties[j]),], col=mypal[1+j], lwd=3)
# }
# # Montezuma County
# points(new.cases.4avg~time.since.first, growing[which(growing$county=="Montezuma"),], pch=3, lwd=2, type="b")
# 
# legend('bottomright', legend=c("all counties","early deaths","Montezuma",gsub(".*\\.","",my.counties)), lwd=c(1,1,2,rep(2, length(my.counties))), pch=c(NA,NA,3,NA,NA), col=c("#AAAAAA66", mypal[1],"black", mypal[2:(length(my.counties)+1)]), ncol=2, cex=.95)
# x <- 1:80
# lines(exp(.2333*x)~x, lty=2)
# lines(exp(.10*x)~x, lty=3)
# legend("topright", legend=c("3 days", "7 days"), lty=c(2,3),title = "Doubling Time", horiz=T, bty="n")




plot(new.cases.4avg~time.since.tenth, growing, type="n", ylab="4 day avg new cases", xlab="days since 10th case", xlim=c(0,40), ylim=c(0,400))
for(i in unique(growing$state.county[which(!growing$state.county %in% bad.counties)])){
  lines(new.cases.4avg~time.since.tenth, growing[which(growing$state.county==i),], col="#AAAAAA66")
}
for(i in unique(growing$state.county[which(growing$state.county %in% bad.counties)])){
  lines(new.cases.4avg~time.since.tenth, growing[which(growing$state.county==i),], col=paste0(mypal[1],"66"))
}
# plot counties of interest
for(j in 1:length(my.counties)){
  lines(new.cases.4avg~time.since.tenth, growing[which(growing$state.county==my.counties[j]),], col=mypal[1+j], lwd=3)
}
# Montezuma County
points(new.cases.4avg~time.since.tenth, growing[which(growing$county=="Montezuma"),], pch=3, lwd=2, type="b")

legend('topleft', legend=c("all counties","early deaths","Montezuma",gsub(".*\\.","",my.counties)), lwd=c(1,1,2,rep(2, length(my.counties))), pch=c(NA,NA,3,NA,NA), col=c("#AAAAAA66", mypal[1],"black", mypal[2:(length(my.counties)+1)]), ncol=2)
x <- 1:80
lines(exp(.2333*x)~x, lty=2)
lines(exp(.10*x)~x, lty=3)
legend("topright", legend=c("3 days", "7 days"), lty=c(2,3),title = "Doubling Time", horiz=T, bty="n")

```


## 'THE CURVE' 
The epidemiological 'curve' of new cases that you've no doubt seen everywhere, the thing we're trying to flatten

``` {r curves, warning=FALSE, echo=FALSE}

plot(new.cases.4avg~date, growing, type="n",log="y", ylab="4 day avg new cases")
for(i in unique(growing$state.county[which(!growing$state.county %in% bad.counties)])){
  lines(new.cases.4avg~date, growing[which(growing$state.county==i),], col="#AAAAAA66")
}
for(i in unique(growing$state.county[which(growing$state.county %in% bad.counties)])){
  lines(new.cases.4avg~date, growing[which(growing$state.county==i),], col=paste0(mypal[1],"66"))
}
# plot counties of interest
for(j in 1:length(my.counties)){
  lines(new.cases.4avg~date, growing[which(growing$state.county==my.counties[j]),], col=mypal[1+j], lwd=3)
}
# Montezuma County
points(new.cases.4avg~date, growing[which(growing$county=="Montezuma"),], pch=3, lwd=2, type="b")

legend('topleft', legend=c("all counties","early deaths","Montezuma",gsub(".*\\.","",my.counties)), lwd=c(1,1,2,rep(2, length(my.counties))), pch=c(NA,NA,3,NA,NA), col=c("#AAAAAA66", mypal[1],"black", mypal[2:(length(my.counties)+1)]))
```


This is real noisy, because in most counties we don't have a ton of new cases per day yet. But one could potentially take heart on the shape of some of these curves. Things *might* be going down (though the accuracy of recent testing data is highly dubious until we're a few days out due to backlogs and such).



**Better 'Curve' assessment?**

``` {r cumulative plot, warning=F, echo=F}


plot(new.cases.4avg~cases, growing, type="n", log="xy",xlab="total cases", ylab="new cases (4 day avg)", yaxt="n", xaxt="n")
axis(2, at = c(1,10,100,1000), labels = c(1,10,100,1000))
axis(1, at= c(1,2,10,20,100,200,1000,2000,10000), labels=c(1,2,10,20,100,200,1000,2000,10000))
for(i in unique(growing$state.county[which(!growing$state.county %in% bad.counties)])){
  lines(new.cases.4avg~cases, growing[which(growing$state.county==i),], col="#AAAAAA66")
}
for(i in unique(growing$state.county[which(growing$state.county %in% bad.counties)])){
  lines(new.cases.4avg~cases, growing[which(growing$state.county==i),], col=paste0(mypal[1],"66"))
}
# plot counties of interest
for(j in 1:length(my.counties)){
  lines(new.cases.4avg~cases, growing[which(growing$state.county==my.counties[j]),], col=mypal[1+j], lwd=3)
}
# Montezuma County
points(new.cases.4avg~cases, growing[which(growing$county=="Montezuma"),], pch=3, lwd=2, type="b")

legend('bottomright', legend=c("all counties","early deaths","Montezuma",gsub(".*\\.","",my.counties)), lwd=c(1,1,2,rep(2, length(my.counties))), pch=c(NA,NA,3,NA,NA), col=c("#AAAAAA66", mypal[1],"black", mypal[2:(length(my.counties)+1)]))
```

This is another way of looking to see when exponential growth gets shut down. X axis is total number of cases, y axis is new cases per day (averaged over past 4 days to get rid of some noise, you can still see weekend slumps in test returns though). When new cases stop being a funciton of total cases (i.e. exponential growth stops), the lines will fall off the diagonal. Unfortunately, no one is really doing that yet. 

```{r deaths vs cummulative deaths, echo=F, warning=F}
### Plotting deaths instead of cases
plot(new.deaths.4avg~deaths, growing, type="n", log="xy",xlab="total deaths", ylab="new deaths (4 day avg)")
#axis(2, at = c(1,10,100,1000), labels = c(1,10,100,1000))
#axis(1, at= c(1,2,10,20,100,200,1000,2000,10000), labels=c(1,2,10,20,100,200,1000,2000,10000))
for(i in unique(growing$state.county[which(!growing$state.county %in% bad.counties)])){
  lines(new.deaths.4avg~deaths, growing[which(growing$state.county==i),], col="#AAAAAA66")
}
for(i in unique(growing$state.county[which(growing$state.county %in% bad.counties)])){
  lines(new.deaths.4avg~deaths, growing[which(growing$state.county==i),], col=paste0(mypal[1],"66"))
}
# plot counties of interest
for(j in 1:length(my.counties)){
  lines(new.deaths.4avg~deaths, growing[which(growing$state.county==my.counties[j]),], col=mypal[1+j], lwd=3)
}
# Montezuma County
points(new.deaths.4avg~deaths, growing[which(growing$county=="Montezuma"),], pch=3, lwd=2, type="b")

legend('bottomright', legend=c("all counties","early deaths","Montezuma",gsub(".*\\.","",my.counties)), lwd=c(1,1,2,rep(2, length(my.counties))), pch=c(NA,NA,3,NA,NA), col=c("#AAAAAA66", mypal[1],"black", mypal[2:(length(my.counties)+1)]))
```

Doing the same thing for deaths rather than cases, we see that, thankfully, we don't have enough county-level deaths for this to work in most cases.

*****

------
### TRACK YOUR STATE
And last but not least, we can look at state totals just for fun (though the NYT does a better job visualizing this)
```{r states, include=F}
my.states <- c("California","Utah","Colorado", "Arizona","Florida", "New York")
state <- read.csv(paste0(gsub(pattern = "/covid-analysis",replacement="", getwd()),"/covid-19-data/us-states.csv"))
state$date <- base::as.Date(state$date)
#state$deaths[which(state$state=="New Jersey" & state$date=="2020-06-25")] # NY added 'probable deaths'
tmp <- state %>% group_by(state) %>% mutate(
  new.cases = c(NA,diff(cases, differences=1))
  , new.deaths = c(NA,diff(deaths, differences=1))
  , min.cases = min(cases[which(new.cases>0)], na.rm=T)
  , min.deaths=min(deaths[which(new.deaths>0)],na.rm=T)
  , first.case = min(date[which(new.cases>0)], na.rm=T)
  , first.death = min(date[which(new.deaths>0)], na.rm=T)
  , tenth.case = min(date[which(cases>=10)], na.rm=T)
  , tenth.death = min(date[which(deaths>=10)], na.rm=T)
  , max.cases=max(cases)
  , max.deaths=max(deaths)
  , new.cases.4avg =rollmean(new.cases, 4, fill=NA, align="right")
  , new.deaths.4avg = rollmean(new.deaths,4, fill=NA, align="right")) %>% arrange(state)
tmp$min.cases[which(tmp$min.cases==Inf)] <- NA
tmp$min.deaths[which(tmp$min.deaths==Inf)] <- NA
tmp$first.case[which(tmp$first.case==Inf)] <- NA
tmp$first.death[which(tmp$first.death==Inf)] <- NA

state.growing <- tmp[which(tmp$date>=tmp$first.case),]
state.growing$time.since.first <- state.growing$date - state.growing$first.case
state.growing$cases.since.min <- state.growing$cases - state.growing$min.cases
state.growing$time.since.tenth <- state.growing$date - state.growing$tenth.case
state.growing$time.since.tenth[which(state.growing$time.since.tenth<0)] <- NA


state.cap <- read.csv(file = "./demog_data/Kaiser_HospitalCapacity.csv") %>% arrange(Location)
state.cap$population <- with(state.cap, Total_Beds *10000/Beds_per_10k)

state.growing$population <- state.cap$population[match(state.growing$state, state.cap$Location)]
state.growing$Beds <- state.cap$Total_Beds[match(state.growing$state, state.cap$Location)]
state.growing$Beds_p10k <- state.cap$Beds_per_10k[match(state.growing$state, state.cap$Location)]
state.growing$ICU <- state.cap$ICU_Beds[match(state.growing$state, state.cap$Location)]
state.growing$ICU_p10k <- state.cap$ICU_Beds_per_10k[match(state.growing$state, state.cap$Location)]
state.growing$cases_p10k <- state.growing$cases/state.growing$population*10000
state.growing$new.cases.4avg_p10k <- state.growing$new.cases.4avg/state.growing$population*10000
state.growing$deaths_p10k <- state.growing$deaths/state.growing$population*10000
state.growing$new.deaths.4avg_p10k <- state.growing$new.deaths.4avg/state.growing$population*10000
state.growing$new.cases.4avg_pICU <- state.growing$new.cases.4avg/state.growing$ICU


CA <- state.growing[which(state.growing$state=="California"),]
BA.c <- growing[which(growing$county %in% c("Santa Clara","San Mateo","San Francisco","Alameda","Contra Costa","Marin")),]
BA <- BA.c %>% group_by(date) %>% summarise(cases = sum(cases, na.rm=T),
                                            deaths = sum(deaths, na.rm=T),
                                            new.cases.4avg =sum(new.cases.4avg, na.rm=T),
                                            new.deaths.4avg = sum(new.deaths.4avg, na.rm=T))


ba.pop <- 7750000-184000-137000-448000 # estimate bay area pop from wikipedia 7.75mil minus sonoma, solano, santa clara county
BA$cases_p10k <- BA$cases/ba.pop*10000
BA$new.cases.4avg_p10k <- BA$new.cases.4avg/ba.pop*10000
BA$deaths_p10k <- BA$deaths/ba.pop*10000
BA$new.deaths.4avg_p10k <- BA$new.deaths.4avg/ba.pop*10000



ca.county.pop <- read.csv("demog_data/CA_county_pop.csv")
ca.county.pop$county <- str_remove(ca.county.pop$CTYNAME, pattern = " County")
ca.county <- growing[which(growing$state=="California"),]
ca.county$population <- ca.county.pop$Pop[match(ca.county$county, ca.county.pop$county)]



ca.county$cases_p10k <- ca.county$cases/ca.county$population*10000
ca.county$new.cases.4avg_p10k <- ca.county$new.cases.4avg/ca.county$population*10000
ca.county$deaths_p10k <- ca.county$deaths/ca.county$population*10000
ca.county$new.deaths.4avg_p10k <- ca.county$new.deaths.4avg/ca.county$population*10000





statesums <- state.growing[which(state.growing$max.cases>=10),] %>% group_by(state, max.cases, max.deaths) %>% summarise( time.to.50 = min(time.since.first[which(cases>= 50)]), time.to.10 = min(time.since.first[which(cases>= 10)]), time.to.10.deaths = min(time.since.first[which(deaths>= 10)]) )
statesums$population <- state.cap$population[match(statesums$state, state.cap$Location)]
statesums$Beds <- state.cap$Total_Beds[match(statesums$state, state.cap$Location)]
statesums$Beds_p10k <- state.cap$Beds_per_10k[match(statesums$state, state.cap$Location)]
statesums$ICU <- state.cap$ICU_Beds[match(statesums$state, state.cap$Location)]
statesums$ICU_p10k <- state.cap$ICU_Beds_per_10k[match(statesums$state, state.cap$Location)]



# double check populations with other source. looks good
# state.pop <- read.csv(file="demog_data/2018Population_by_age_gender.csv")
# tot.pop <- state.pop %>% filter(Gender=="Total" & Age=="Total" & !LocationDesc %in% c("Puerto Rico", "United States (States and DC)") & YEAR==2017) %>% arrange(LocationAbbr)
# 
# tot.pop$pop1 <- state.cap$pop1[match(tot.pop$LocationDesc, state.cap$Location)]
```


``` {r Bay Area per capita}

quartz(width=6, height=3.5)
jpeg(paste0("BayARea_percaptia_",max(state.growing$date),".jpg"),width = 6, height=3.5, units = "in", res = 600)
par(mfrow=c(1,2), mar=c(3.5,3.5,1,1), mgp=c(2,1,0))
plot(new.cases.4avg~date, CA, type="l", lwd=3, ylab="New Cases (4 day avg)")
lines(new.cases.4avg~date, BA, col="blue", lwd=2)
legend("topleft", legend=c("CA total","Bay Area"), lwd=c(3,2), col=c("black","blue"))
plot(new.cases.4avg_p10k~date, ca.county[which(ca.county$population>80000),], type="n", lwd=3, ylab="New Cases per 10k pop", ylim=c(0,3))
for(i in unique(ca.county$county[which(ca.county$population>80000)])){
  lines(new.cases.4avg_p10k~date, ca.county[which(ca.county$county==i),], col="#66666666")
}
lines(new.cases.4avg_p10k~date, ca.county[which(ca.county$county=="Los Angeles"),], col="red")
for(i in c("Marin","San Francisco","San Mateo","Santa Clara", "Alameda","Contra Costa")){
  lines(new.cases.4avg_p10k~date, ca.county[which(ca.county$county==i),], col="#4DAF4A") #"#377EB866")
}
lines(new.cases.4avg_p10k~date, CA, lwd=3)
lines(new.cases.4avg_p10k~date, BA, col="blue", lwd=3)
legend("topleft", legend=c("all counties","LA County", "Bay Counties"), lwd=1, col=c("#66666666","red","#4DAF4A"), cex=.8)
quartz.save("CA_BayArea_comparison_June20v2.png", dpi=300)


```



``` {r state curves, warning=FALSE, echo=FALSE}

plot(new.cases.4avg~date, state.growing, type="n",log="y", ylab="4 day avg new cases", main="New Cases over time")
for(i in unique(state.growing$state)){
  lines(new.cases.4avg~date, state.growing[which(state.growing$state==i),], col="#AAAAAA66")
}

# plot counties of interest
for(j in 1:length(my.states)){
  lines(new.cases.4avg~date, state.growing[which(state.growing$state==my.states[j]),], col=mypal[1+j], lwd=3)
}

legend('topleft', legend=c("all states",gsub(".*\\.","",my.states)), lwd=c(1,rep(2, length(my.states))), col=c("#AAAAAA66", mypal[2:(length(my.states)+1)]))
```





``` {r state curves per capita, warning=FALSE, echo=FALSE}

plot(new.cases.4avg_p10k~date, state.growing, type="n", ylab="4 day avg new cases", main="New Cases over time")
for(i in unique(state.growing$state)){
  lines(new.cases.4avg_p10k~date, state.growing[which(state.growing$state==i),], col="#AAAAAA66")
}

# plot counties of interest
for(j in 1:length(my.states)){
  lines(new.cases.4avg_p10k~date, state.growing[which(state.growing$state==my.states[j]),], col=mypal[1+j], lwd=3)
}

legend('topleft', legend=c("all states",gsub(".*\\.","",my.states)), lwd=c(1,rep(2, length(my.states))), col=c("#AAAAAA66", mypal[2:(length(my.states)+1)]))
```

``` {r state curves deathes per capita, warning=FALSE, echo=FALSE}

plot(new.deaths.4avg_p10k~date, state.growing, type="n", ylab="4 day avg new cases", main="New Cases over time")
for(i in unique(state.growing$state)){
  lines(new.deaths.4avg_p10k~date, state.growing[which(state.growing$state==i),], col="#AAAAAA66")
}

# plot counties of interest
for(j in 1:length(my.states)){
  lines(new.deaths.4avg_p10k~date, state.growing[which(state.growing$state==my.states[j]),], col=mypal[1+j], lwd=3)
}

legend('topleft', legend=c("all states",gsub(".*\\.","",my.states)), lwd=c(1,rep(2, length(my.states))), col=c("#AAAAAA66", mypal[2:(length(my.states)+1)]))
```
``` {r state curves per ICU bed}

plot(new.cases.4avg_pICU~date, state.growing, type="n", ylab="4 day avg new cases", main="New Cases over time")
for(i in unique(state.growing$state)){
  lines(new.cases.4avg_pICU~date, state.growing[which(state.growing$state==i),], col="#AAAAAA66")
}

# plot counties of interest
for(j in 1:length(my.states)){
  lines(new.cases.4avg_pICU~date, state.growing[which(state.growing$state==my.states[j]),], col=mypal[1+j], lwd=3)
}

legend('topleft', legend=c("all states",gsub(".*\\.","",my.states)), lwd=c(1,rep(2, length(my.states))), col=c("#AAAAAA66", mypal[2:(length(my.states)+1)]))
```


``` {r where are we headed?, warning=FALSE, echo=FALSE}

#quartz(width=6, height=6)
jpeg(paste0("State_percaptia_",max(state.growing$date),".jpg"),width = 6, height=6, units = "in", res = 600)
par(mfrow=c(2,1), mar=c(3,4,1,1))
plot(new.cases.4avg_p10k~date, state.growing, type="n", ylab="Cases per 10k pop", main="New Cases")
for(i in unique(state.growing$state)){
  lines(new.cases.4avg_p10k~date, state.growing[which(state.growing$state==i),], col="#AAAAAA66")
}

# plot counties of interest
for(j in 1:length(my.states)){
  lines(new.cases.4avg_p10k~date, state.growing[which(state.growing$state==my.states[j]),], col=mypal[1+j], lwd=3)
}

legend('topleft', legend=c("all states",gsub(".*\\.","",my.states)), lwd=c(1,rep(2, length(my.states))), col=c("#AAAAAA66", mypal[2:(length(my.states)+1)]))



plot(new.deaths.4avg_p10k~date, state.growing, type="n", ylab="Deaths per 10k pop", main="New Deaths", xlim=c(min(state.growing$date)+21, max(state.growing$date + 21)))
for(i in unique(state.growing$state)){
  lines(new.deaths.4avg_p10k~date, state.growing[which(state.growing$state==i),], col="#AAAAAA66")
}

# plot counties of interest
for(j in 1:length(my.states)){
  lines(new.deaths.4avg_p10k~date, state.growing[which(state.growing$state==my.states[j]),], col=mypal[1+j], lwd=3)
}
text(x=as.Date("2020-07-01"),y=.2, labels="Deaths = 4wk\nlagging indicator")
lines(y=c(.1,.1), x=c(max(state.growing$date), as.Date("2020-07-24")), lwd=3)
arrows(y0=c(.55, .4), x0 = as.Date(c("2020-05-05", "2020-07-02")), y1=c(0.5, 0.5), x1=as.Date(c("2020-04-15","2020-07-24")), lwd=3, length=0.1)
text(x = as.Date(c("2020-05-30", "2020-06-10")),y=c(.55, .4),labels=c("New York in Apr", "Arizona in Aug?"), col=mypal[c(7,5)])
dev.off()
```







**Same curves, but standardizing the time axis to put all counties on the same timeline:**

```{r state the CURVE standardized by first case, echo=F, warning=F}
par(mar=c(4,4,1,1))

plot(new.cases.4avg~time.since.first, state.growing, type="n",log="y", ylab="4 day avg new cases", main="New Cases since First Case")
for(i in unique(state.growing$state)){
  lines(new.cases.4avg~time.since.first, state.growing[which(state.growing$state==i),], col="#AAAAAA66")
}

# plot counties of interest
for(j in 1:length(my.states)){
  lines(new.cases.4avg~time.since.first, state.growing[which(state.growing$state==my.states[j]),], col=mypal[1+j], lwd=3)
}

legend('bottomright', legend=c("all states",gsub(".*\\.","",my.states)), lwd=c(1,rep(2, length(my.states))), col=c("#AAAAAA66", mypal[2:(length(my.states)+1)]))




plot(new.cases.4avg~time.since.tenth, state.growing, type="n",log="y", ylab="4 day avg new cases", main="Total Cases since 10th Case")
for(i in unique(state.growing$state)){
  lines(new.cases.4avg~time.since.tenth, state.growing[which(state.growing$state==i),], col="#AAAAAA66")
}

# plot counties of interest
for(j in 1:length(my.states)){
  lines(new.cases.4avg~time.since.tenth, state.growing[which(state.growing$state==my.states[j]),], col=mypal[1+j], lwd=3)
}

legend('bottomright', legend=c("all states",gsub(".*\\.","",my.states)), lwd=c(1,rep(2, length(my.states))), col=c("#AAAAAA66", mypal[2:(length(my.states)+1)]))

```

**New Cases vs Total Cases**

``` {r state cumulative plot, warning=F, echo=F}
plot(new.deaths.4avg~deaths, state.growing, type="n",log="xy", ylab="4 day avg new deaths")
for(i in unique(state.growing$state)){
  lines(new.deaths.4avg~deaths, state.growing[which(state.growing$state==i),], col="#AAAAAA66")
}

# plot counties of interest
for(j in 1:length(my.states)){
  lines(new.deaths.4avg~deaths, state.growing[which(state.growing$state==my.states[j]),], col=mypal[1+j], lwd=3)
}

legend('topleft', legend=c("all states",gsub(".*\\.","",my.states)), lwd=c(1,rep(2, length(my.states))), col=c("#AAAAAA66", mypal[2:(length(my.states)+1)]))

```


No states really seem to be peaking and falling off of the exponential growth phase yet... 







``` {r totalsm, echo=F}
# plot(max.cases~max.deaths, countysums[which(countysums$state %in% c("Colorado","Utah","New Mexico", "Arizona")),], pch=16, col=state)
# 
# tmp <- countysums[which(countysums$state %in% c("Colorado","Utah","New Mexico", "Arizona")),] %>% arrange(max.cases)
# tmp[which(tmp$state.county %in% my.counties),]
```
